Query Optimization for Dynamic Imputation

### 2017-05-11 paper/algorithm.tex
Update Imputation Strategies section

Scan costs are estimated based on the number of tuples, page size, and a unit-less I/O cost per page. 
We assume that I/O costs dominate CPU costs for relational operators. 
A database implementation for which this assumption is not true, e.g. an in memory database, will need a different cost model.
Join costs are estimated as a function of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead so we assume negligible time costs for those operations.

We extend these heuristics to account for the new operators: \textit{Drop} and \textit{Impute}.
\textit{Drop} is a special case of a filter, so its time complexity is correspondingly negligible. 
As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.

\noindent{\textbf{Constructing $T$}}: The choice of $T$ depends on the imputation method in use, and as with $P$, there are several intuitions which guide the choice of $T$.
First, $\textsc{Time}$ is computed in terms of the I/O cost per page so $T$ must return a value in these units.
It can be unnatural to compute the cost of imputation in terms of the I/O cost, so it may be simpler to compute the per-tuple cost and then convert.
Second, $T$ should \emph{increase} as the number of tuples passed to the imputation function increases and it should be proportional to the complexity of the imputation algorithm.
As with $P$, the ordering between queries generated by $T$ is more important than the precision of its estimates.
\todo{Micah, how did you deal with the IO/CPU cost problem?}

@@

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by adding
time and penalty functions (\Cref{sec:cost-model}) which describe the new algorithm.
For our purposes, an imputation strategy is a function $I(R, C) \rightarrow R'$ which fills in the missing
attributes $C$ in a relation $R$, returning a new relation $R'$.
The flexibility of this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation.

\subsubsection{Regression trees}

New imputation algorithms can also be added. As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained-equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision trees algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation~\cite{witten2016data}.

\[P(q) = \frac{1}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}\]

The reference algorithm used by the \textit{Impute} operator proceeds by iteratively fitting
decision trees for the target imputation attribute on a subset of the data. In each
iteration, the missing values of a separate attribute are replaced with newly imputed values.
In one {\it epoch} of the algorithm, each of the target imputation
attributes is filled in with imputed values in a single iteration.
With each epoch, the quality of the imputation improves as values progressively
reflect more accurate relationships between attributes. The algorithm terminates when
convergence is achieved (i.e.\ the imputed values do not change across epochs) or a fixed
number of epochs is reached.

These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
multiple imputation could incorporated as another avenue of assessing imputation confidence.

>>>>

Scan costs are estimated based on the number of tuples, page size, and a unit-less I/O cost per page. 
We assume that I/O costs dominate CPU costs for relational operators. 
A database implementation for which this assumption is not true, e.g. an in-memory database, will need a different cost model.
Join costs are estimated as a function of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead so we assume negligible time costs for those operations.

We extend these heuristics to account for the new operators: \textit{Drop} and \textit{Impute}.
\textit{Drop} is a special case of a filter, so its time complexity is correspondingly negligible.

\noindent{\textbf{Constructing $T$}}: As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on
the properties of the underlying algorithm via a function $T$. If we assume
that the \textit{Impute} routine is CPU-bound, then we will adjust
\textsc{Time} by some constant factor $f$, thereby scaling the result to be
comparable with IO-bound routines.

The form of $T$ depends on the specific imputation method, and as with $P$,
there are some principles to capture the intuition.  $T$ should \emph{increase}
as the number of tuples and complete and dirty attributes passed to the
imputation function increases and it should be take into account the complexity
of the imputation algorithm. Analysis of the time complexity of different
imputation algorithms, especially with in-database execution, is beyond the
scope of this paper. As with $P$, it is the ordering between queries
generated by $T$ that is more important than the precision of estimates.

@@

\ProjectName{} is designed so that any imputation strategy
can be plugged in with minimal effort and without changes to the optimizer, simply by supplying
penalty and time functions (\Cref{sec:cost-model}) pertinent to the algorithm.
These can then be queried by the optimizer during planning.  The flexibility of
this approach aligns with the spirit of query optimization, which aims to
reconcile performance with a declarative interface to data manipulation.

\subsubsection{Decision trees}

New imputation algorithms can also be added. As a reference implementation,
\ProjectName{} uses a general-purpose imputation strategy based on
chained-equation decision trees.  Chained-equation imputation
methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by
iteratively fitting a predictive model of one missing attribute, conditional on
all complete attributes and all other missing attributes. Chained-equation
decision trees algorithms are found to be effective~\cite{akande2015empirical}
and are widely used in epidemiological domains~\cite{burgette2010multiple}.

These imputation algorithms are commonly used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset~
\cite{gelman2006data}. In \ProjectName{}, we consider only single imputation, though
multiple imputation could be incorporated as another avenue of assessing imputation confidence.

In the reference implementation, the chained-equation algorithm proceeds by
iteratively fitting decision trees for the target imputation attributes. In
each iteration, the missing values of a separate attribute are replaced with
newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved. (That is, when the imputed
values of one column are not updated by subsequent iterations that update that
column.) For our experiments, we use Weka's \texttt{REPTree}
implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

\begin{equation*}
    \begin{aligned}
    T(q) = k * &|\textsc{Dirty}(q)| * \textsc{Card}(q) * \\
               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q))
    \end{aligned}
\end{equation*}

Separately, we specify a heuristic penalty that takes into account the
improvement in imputation quality from more tuples and attributes as well as
the diminishing returns to more data.

\[
    P(q) = \frac{1}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
\]
